---
title: "Assignment 10: Data Scraping"
author: "Flannery McNair"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(rvest)
library(tidyverse)
library(lubridate)

getwd()

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
website <- read_html(
  "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024")
the_website <- read_html('https://www.eia.gov/electricity/state/')
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
water_system_name <- website %>% 
  html_node('table:nth-child(7)') %>%
  html_node('td:nth-child(2)') %>% 
  html_text()

pwsid <- website %>% 
  html_node('td:nth-child(5)') %>% 
  html_text()

ownership <- website %>% 
  html_node('td:nth-child(4)') %>% 
  html_text()

water_system_name
pwsid
ownership

```

```{r}
MGD_1 <- website %>% 
  html_nodes('table:nth-child(30)') %>% 
  html_nodes('td:nth-child(3)') %>%
  html_text()

MGD_2 <- website %>% 
  html_nodes('table:nth-child(30)') %>% 
  html_nodes('td:nth-child(6)') %>%
  html_text()

MGD_3 <- website %>% 
  html_nodes('table:nth-child(30)') %>% 
  html_nodes('td:nth-child(9)') %>%
  html_text()

mgd_total <- c(MGD_1, MGD_2, MGD_3)
mgd_total
```

4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 
months <- c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct",
            "Nov","Dec")
month_num <- match(substr(months, 1, 3), month.abb)


mgd_data <- data.frame(
  "Water System Name" = rep(water_system_name, 12),
  "PWSID" = rep(pwsid, 12), 
  "Ownership" = rep(ownership,12), 
  "Month" = as.character(months),
  "MGD" = as.numeric(gsub(',','',mgd_total)),
  "Date" = as.Date(sprintf("2024-%02d-01", month_num))
)

mgd_data

#5 
ggplot(mgd_data, aes(x = Date, y = MGD)) +
  geom_line() +
  geom_point() +
  labs(
    title = paste("Maximum Daily Withdrawals for", water_system_name, "— 2024"),
    x = "Month",
    y = "Max Day Use (MGD)"
  ) +
  theme_classic()

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs - "PWSID" and "year" - that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. 
scrape_func <- function(the_year, the_id) {
  the_url <- ifelse(
    the_year>=2024, 
    'https://www.ncwater.org/WUDC/app/LWSP/', 
    paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=',the_id,'&year=', the_year)
  )
  the_website <- read_html(the_url)
  
  water_system_name <- the_website %>% 
    html_node('table:nth-child(7)') %>%
    html_node('td:nth-child(2)') %>% html_text()

  pwsid <- the_website %>% 
    html_node('td:nth-child(5)') %>% 
    html_text()

  ownership <- the_website %>% 
    html_node('td:nth-child(4)') %>% 
    html_text()
  
  MGD_1 <- the_website %>% 
    html_nodes('table:nth-child(31)') %>% 
    html_nodes('td:nth-child(3)') %>%
    html_text()

  MGD_2 <- the_website %>% 
    html_nodes('table:nth-child(31)') %>% 
    html_nodes('td:nth-child(6)') %>%
    html_text()

  MGD_3 <- the_website %>% 
    html_nodes('table:nth-child(31)') %>% 
    html_nodes('td:nth-child(9)') %>%
    html_text()

  mgd_total <- c(MGD_1, MGD_2, MGD_3)
  
  months <- month.abb
  month_num <- match(substr(months, 1, 3), month.abb)
  
  df <- data.frame(
    "Water System Name" = rep(water_system_name, 12),
    "PWSID" = rep(pwsid, 12), 
    "Ownership" = rep(ownership,12), 
    "Month" = as.character(months),
    "MGD" = as.numeric(gsub(',','',mgd_total)),
    "Date" = as.Date(sprintf("%04d-%02d-01", the_year, month_num))
    )
}



```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 

durham_2020 <- scrape_func(2020, '03-32-010')

str(durham_2020)

```

8. Use the function above to extract data for Cary (PWSID = '03-32-010') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 
scrape_func_2 <- function(the_year) {
  the_url <- ifelse(
    the_year>=2024, 
    'https://www.ncwater.org/WUDC/app/LWSP/', 
    paste0(
      'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-92-020&year=',
      the_year)
  )
  the_website <- read_html(the_url)
  water_system_name <- the_website %>% 
    html_node('table:nth-child(7)') %>%
    html_node('td:nth-child(2)') %>% html_text()

  pwsid <- the_website %>% 
    html_node('td:nth-child(5)') %>% 
    html_text()

  ownership <- the_website %>% 
    html_node('td:nth-child(4)') %>% 
    html_text()
  
  MGD_1 <- the_website %>% 
    html_nodes('table:nth-child(32)') %>% 
    html_nodes('td:nth-child(3)') %>%
    html_text()

  MGD_2 <- the_website %>% 
    html_nodes('table:nth-child(32)') %>% 
    html_nodes('td:nth-child(6)') %>%
    html_text()

  MGD_3 <- the_website %>% 
    html_nodes('table:nth-child(32)') %>% 
    html_nodes('td:nth-child(9)') %>%
    html_text()

  mgd_total <- c(MGD_1, MGD_2, MGD_3)
  
  months <- month.abb
  month_num <- match(substr(months, 1, 3), month.abb)
  
  df <- data.frame(
    "Water System Name" = rep(water_system_name, 12),
    "PWSID" = rep(pwsid, 12), 
    "Ownership" = rep(ownership,12), 
    "Month" = as.character(months),
    "MGD" = as.numeric(gsub(',','',mgd_total)),
    "Date" = as.Date(sprintf("%04d-%02d-01", the_year, month_num))
    )
}

cary_2020 <- scrape_func_2(2020)


```

9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 

cary_18_23 <- seq(2018,2023) %>% map(scrape_func_2) %>% bind_rows()

cary_18_23_plot <- ggplot(
  cary_18_23, aes(x = month(Date), y = MGD, group = year(Date),
                  color = factor(year(Date)))) +
  geom_line() +
  geom_point() +
  geom_smooth(aes(group = 1), method = "loess", color = "black") +
  scale_x_continuous(breaks = 1:12,labels = month.abb) +
  labs(
    title = "Maximum Daily Withdrawals by Month in Cary (2018–2023)",
    x = "Month",
    y = "Max Daily Use (MGD)",
    color = "Year"
  ) +
  theme_minimal()

cary_18_23_plot
```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer: Yes, there is much more water usage in the summer, and it has been slightly more in the recent years but not significantly. 
>
